---
title: "Speed&Distance Predictive_Analysis"
author: "Akolade Sofiyyah Iwalewa,22201441"
date: "2022-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to data set
The data gives the speed of cars (mph) and the distances (ft) taken to stop. The cars data set that comes with R by default

```{r}
# Having an overview of the data set
head(cars)
```

```{r}
# Checking the dimension of the data
dim(cars)
```

This data set consists of 50 observations (rows) and 2 variables(columns) – distance and speed.

```{r}
# creating a scatter plot to Visualise the linear relationship between the predictor and the response variable
pairs(cars)
```

There exist a positive relationship between the distance and speed.

```{r}
# Calculating the correlation co-efficient that measures the strength of the linear relationship between the predictor and response variable.
cor(cars)
```
0.80 indicates a strong linear relationship between distance and speed.

```{r}
# Getting the distribution of the variables by numerical values.
summary(cars)
```

```{r}
# Getting the distribution of the variables by plotting graphs.
par(mfrow=c(2, 2))
boxplot(cars$speed, main="Speed")
boxplot(cars$dist, main="Distance")
plot(density(cars$speed), main="Density Plot: Speed")
polygon(density(cars$speed), col="red")
plot(density(cars$dist), main="Density Plot: Distance")
polygon(density(cars$dist), col="red")
```

The minimum speed recorded was 4 km/h and the maximum speed is 25 km/h. 25% of the car speed fall bellow 12 km/h, 50% of the car speed fall below 15 km/h,75% of the car speed fall below 19 km/h. The minimum distance recorded was 2 km and the maximum distance was 120 km. 25% of the car distance fall below 26 km , 50% of the car distance fall below 36km, 75% of the car distance fall below 56 km. From the box plot, a particular seem to an outlier.

```{r}
# Checking the distance that is greater 100
which(cars$dist>100)
cars[49,]
```

Car 49 has a distance of 120 km  which is greater than 100

I will be fitting the linear model below: distancei = β0 + β1speedi + i


```{r}
# Fitting the linear model.
SXX = sum((cars$speed-mean(cars$speed))^2)
SXY = sum((cars$speed-mean(cars$speed))
*(cars$dist-mean(cars$dist)))
beta1 <- SXY / SXX
beta0 <- mean(cars$dist) - beta1 * mean(cars$speed)
c(beta0,beta1)
```

```{r}
linearMod <- lm(dist ~speed, data=cars)
linearMod
```

THe model becomes: distance ˆi = −17.58 + 3.93 × speed

The baseline of 0 mph actually has a negative stopping distance −17.58. Thus we restrict the interpretability of the model to speeds 4.48 (≈ 17.58/3.93) or more mph. This model estimates that increasing the speed by 1 mph will result in an extra 3.93 feet of stopping distance. to aid interpretation of the intercept we can minus the mean
from the predictor variable speed. This will be 0 when the speedi is at the average speed which is X¯ = 15.4


```{r}
# Substracting the mean from the predictor variable speed
cars$cen_speed <-cars$speed-mean(cars$speed)
linearMod <- lm(dist ~cen_speed, data=cars)
linearMod
```
Thus I have my new model to be: distance ˆi = 42.98 + 3.93 × (speedi − average(speed))


```{r}
# Creating a line chart for speed and distance
plot(cars$speed,cars$dist)
lines(cars$speed,fitted(linearMod),col="red")
```

Assumptions of the LS estimators: I will be testing the LS estimators assumptions.

1. It is required there is  zero conditional mean and constant variance (constant variability about the zero mean

```{r}
# Getting the summary of the residuals
summary(residuals(linearMod))
```

```{r}
#Creating a scatter plot for the residuals
plot(cars$speed,residuals(linearMod))
```

Note: To perform model inference (T-test, F-test, CI, PI) the errors must be normally distributed or at least approximately normally distributed

```{r}
# Creating a box-plot and density plot for the residuals
par(mfrow=c(2, 1))
boxplot(residuals(linearMod), main="Residuals")
plot(density(residuals(linearMod)),
main="Density Plot: Residuals")
polygon(density(residuals(linearMod)), col="red")
```

```{r}
# Creating a QQ plot
qqnorm(residuals(linearMod),main="QQ plot",pch=19)
qqline(residuals(linearMod))
```

I will be using the Shapiro-Wilk normality test to test for nomality.

```{r}
residuals(linearMod)
```

The p-value 0.02 < 0.05 implying that the distribution of the data is significantly different from a normal distribution. Hence, we cannot assume normality.


Cooks distance: I will be using the Cooks distance to compute the influence exerted by each data point on the predicted outcome.

```{r}
cooksd <- cooks.distance(linearMod)
plot(cooksd, pch="*", cex=2, main="Influential Obs
by Cooks distance")
# add cutoff line
abline(h = 4*mean(cooksd, na.rm=T), col="red")
# add labels
text(x=1:length(cooksd)+1, y=cooksd,
labels=ifelse(cooksd>4*mean(cooksd,
na.rm=T),names(cooksd),""), col="red")
```

I will remove the Influential observation (it is likely that it is a typo) a speed of 24 has a typical stopping distance of 40 feet and a distance of 120 feet typically corresponds to a speed of 40

```{r}
cars[49,]
cars = cars[-49,]
linearMod <- lm(dist ~cen_speed, data=cars)
linearMod
```

```{r}
# Check for normaility
shapiro.test(residuals(linearMod))
residuals(linearMod)
```

```{r}
qqnorm(residuals(linearMod),main="QQ plot",pch=19)
qqline(residuals(linearMod))
```

The p-value 0.08 > 0.05 implying that the distribution of the data is not significantly different from a normal distribution. In other words, we can assume normality.


Hypothesis testing


```{r}
N = length(cars$cen_speed)
MSE = sum(linearMod$residuals^2/(N-2))
SXX = sum((cars$cen_speed-mean(cars$cen_speed))^2)
VARB0 = MSE*(1/ N + (mean(cars$cen_speed)^{2}/SXX))
T = (linearMod$coefficients[1]-0)/sqrt(VARB0)
```

```{r}
alpha = 0.05
TDIST = qt(1-alpha/2, N-2)
PVALUE = 2 *( 1- pt(T, df = N- 2))
```

At the 5% level of significance, the evidence is not strong enough to indicate that β0 = 0. Indicating that when the speed is at 15.4 (the mean) the stopping distance is non-zero.

```{r}
N = length(cars$cen_speed)
MSE = sum(linearMod$residuals^2/(N-2))
SXX = sum((cars$cen_speed-mean(cars$cen_speed))^2)
VARB1 = MSE/SXX
T = (linearMod$coefficients[2]-0)/sqrt(VARB1)
```

```{r}
alpha = 0.05
TDIST = qt(1-alpha/2, N-2)
PVALUE = 2*(1-pt(T, df = N - 2))
```

At the 5% level of significance, the evidence is not strong enough to indicate that β1 = 0. Indicating that a relation exists between speed and stopping distance.

```{r}
MSR = sum((fitted(linearMod) - mean(cars$dist))^2) / 1
MSE = sum(linearMod$residuals^2/(N-2))
F = MSR/MSE 
alpha = 0.05
FDIST = qf(1-alpha,1,N-2) 
PVALUE = pf(1-F, 1, N - 2) 
```

At the 5% level of significance, the evidence is not strong enough to indicate that β1 = 0. Indicating that a relation exists between speed and stopping distance.


```{r}
N = length(cars$cen_speed)
MSE = sum(linearMod$residuals^2/(N-2))
SXX = sum((cars$cen_speed-mean(cars$cen_speed))^2)
VARB0 = MSE*(1/ N + (mean(cars$cen_speed)^{2}/SXX))
alpha=0.05
beta0 = linearMod$coefficients[1]
c(beta0 - qt(1-alpha/2,N-2)*sqrt(VARB0),
beta0 + qt(1-alpha/2,N-2)*sqrt(VARB0))
```

We are 95% confident that β0 lies between 38.0 < β0 < 46.1

```{r}
N = length(cars$cen_speed)
SSE = sum(linearMod$residuals^2)
MSE = SSE/(N-2)
SXX = sum((cars$cen_speed - mean(cars$cen_speed))^2)
VARB1 = MSE/SXX
beta1= linearMod$coefficients[2]
alpha=0.05
c(beta1 - qt(1-alpha/2,N-2)*sqrt( VARB1),
beta1 + qt(1-alpha/2,N-2)*sqrt( VARB1))
```

We are 95% confident that β1 representing the average increase in stopping distance given a one unit increase in speed is between 2.85 and 4.43 feet.

```{r}
SST = sum((cars$dist-mean(cars$dist))^2)
SSE = sum(linearMod$residuals^2)
R2 <- (SST - SSE) /SST 
R2
```
Approximately 64% of the observed variation in stopping distances can be explained by the cars speed.

```{r}
N = length(cars$cen_speed)
RMSE = sqrt(SSE/(N-2))
RMSE
```

So we can say that the speed accurately predicts distance with about 14 feet error on average.

```{r}
N = length(cars$cen_speed)
SXX = sum((cars$cen_speed - mean(cars$cen_speed))^2)
MSE = SSE/(N-2)
VAR_Y = MSE*(1/N+(cars$cen_speed-mean(cars$cen_speed))^2/SXX)
Yhat = fitted(linearMod)
cbind(Yhat- qt(1-alpha/2,N-2)*sqrt(VAR_Y),
Yhat + qt(1-alpha/2,N-2)*sqrt(VAR_Y))
plot(cars$speed,cars$dist,xlab="Speed",ylab="Distance")
lines(cars$speed,Yhat,col="blue")
lines(cars$speed,Yhat+qt(1-alpha/2,N-2)*sqrt(VAR_Y),col="red")
```

```{r}
N = length(cars$cen_speed)
SXX = sum((cars$cen_speed - mean(cars$cen_speed))^2)
MSE = SSE/(N-2)
Var_E = MSE*(1 + 1/N + (cars$cen_speed-mean(cars$cen_speed))^2/SXX)
Yhat = fitted(linearMod)
cbind(Yhat- qt(1-alpha/2,N-2)*sqrt( Var_E),
Yhat + qt(1-alpha/2,N-2)*sqrt( Var_E))
plot(cars$speed,cars$dist,xlab="Speed",ylab="Distance")
lines(cars$speed,Yhat,col="blue")
lines(cars$speed,Yhat +qt(1-alpha/2,N-2)*sqrt(Var_E),col="red")
lines(cars$speed,Yhat -qt(1-alpha/2,N-2)*sqrt(Var_E),col="red")
lines(cars$speed,Yhat-qt(1-alpha/2,N)*sqrt(VAR_Y),col="red")
```